How to select predictive models for decision-making or causal inference?

### 2023-11-19 how_to_select_causal_models.tex
iter condense

Progress in machine learning brings predictive models to new health
data \cite{beam2018big,rajkomar2019machine}, with the promise of
precision medicine.
Automated analysis of
medical images is increasingly accurate, \emph{eg} for brain images,
\cite{khojaste2022deep,zhang2019radiological} or mammography
\cite{yala2019deep,shen2019deep,nassif2022breast}.
New prognostic models leverage routinely-collected
patient records
\cite{mooney2018bigdata}: predicting heart
failure from claims \cite{desai2020comparison}, suicide attempts
from questionnaires \cite{simon2018predicting}...
Clinical notes contain much prognostic information
but require text modeling  \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
Data may be difficult to
control and model, but the accuracy of the prediction can be verified
on left-out data
\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
Given a model predicting a health outcome, precision medicine
would like it to guide decisions: will an individual benefit from an
intervention such as surgery \cite{fontana2019can}?
Contrasting predictions with and without the treatment gives an answer,
but statistical validity requires causal inference
\cite{snowden_implementation_2011,blakely2020reflection}.

@@

There is a rich causal-inference statistical literature, which bridges to
predictive modeling via so-called \emph{outcome models}, or G-computation,
G-formula \cite{robins_role_1986}, Q-model \cite{snowden_implementation_2011},
conditional mean regression \cite{wendling_comparing_2018}. A central challenge
of inference of treatment effects is that of confounding: spurious associations
between treatment allocation and baseline health, \emph{eg} only prescribing a
drug to mild cases \citep{hernan_causal_2020,vanderweele2019principles}.
Controlled allocation of treatment, as in randomized controlled trials (RCTs),
alleviate this concern. Yet most machine-learning models are trained on
\emph{observational} data, close to real-word practice
\cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Causal inference has been central to epidemiology, typically with methods that
model treatment assignment \cite{austin_moving_2015,grose_use_2020}, based on
propensity scores \cite{rosenbaum_central_1983}. Recent empirical results
\cite{wendling_comparing_2018,dorie_automated_2019} show benefits of outcome
modeling to estimate average treatment effects. Maybe a greater benefit is
that these methods naturally go beyond average effects, estimating
individualized or conditional average treatment effects (CATE), central to
precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

%\idea{There is a variety of models for g-estimation}
There is an explosion of outcome modeling or machine learning methods.
For instance, many deep-learning methods have been developed for medical
image analysis \cite{shen2017deep,monshi2020deep}.
Even outcome-modeling methods specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand.
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models (see example from \cite{dorie_automated_2019} developed in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

Standard practices to select models in predictive settings rely on
the error on the outcome
\cite{poldrack2020establishment,varoquaux2022evaluating}. However, as we
will see, these practices may not pick the best models
for causal inference, as they can be misled by inhomogeneities due to
treatment allocation.
%
Given complex, potentially noisy, data, which model is to be most trusted to
yield valid causal estimates? As no single learner performs
best on all data sets, there is a pressing need for clear guidelines to select
outcome models for causal inference.

@@

In this paper, we study \textit{model selection procedures}
in practical settings: \textit{finite samples} settings and without
\textit{well-specification} assumption. Asymptotic causal-inference
theory calls for complex risks, but a practical question is
whether model-selection procedures, that rely on data split, can estimate
these risks reliably enough. Indeed, they
come with more quantities to estimate, which may
bring additional variance, leading to worse model selection.

We first illustrate the problem of causal model
selection and briefly review prior art. Then the methods
\autoref{sec:framework} sets causal model selection in the
\emph{potential outcome} framework and details the causal risks and
model-selection procedure. The next \autoref{sec:theory} gives theoretical
results. The results \autoref{sec:empirical_study} details a thorough empirical
study, covering many different settings. Finally, we
discuss the findings in the last \autoref{sec:discussion}.
%
Results outline how to best select outcome models for causal
inference with an adapted
cross-validation to estimate the so-called $R\text{-risk}$.
This risk compensates for systematic
differences between treated and non-treated individuals using
two \emph{nuisance} models,
themselves estimated from data and thus imperfect; yet these
imperfections do not undermine the $R\text{-risk}$.

>>>>

Progress in machine learning brings predictive models to new health data
\cite{beam2018big,rajkomar2019machine}, with the promise of precision medicine.
Predictive models are increaslingly used for image analysis
\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
routinely-collected patient records
\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}. Data may
be difficult to control and model, but the accuracy of the prediction can be
verified on left-out data
\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
Given a model predicting a health outcome, precision medicine would like it to
guide decisions: will an individual benefit from an intervention such as surgery
\cite{fontana2019can}? Contrasting predictions with and without the treatment
gives an answer, but statistical validity requires causal inference
\cite{snowden_implementation_2011,blakely2020reflection}.

@@

There is a rich causal-inference statistical literature, which bridges to
predictive modeling via so-called \emph{outcome models}, or G-computation,
G-formula \cite{robins_role_1986}, Q-model \cite{snowden_implementation_2011},
conditional mean regression \cite{wendling_comparing_2018}. A central challenge
of inference of treatment effects is that of confounding: spurious associations
between treatment allocation and baseline health, \emph{eg} only prescribing a
drug to mild cases \citep{hernan_causal_2020,vanderweele2019principles}.
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Causal inference has been central to epidemiology, typically with methods that
model treatment assignment \cite{austin_moving_2015,grose_use_2020}, based on
propensity scores \cite{rosenbaum_central_1983}.  However, outcome modeling go
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

% Recent empirical results \cite{wendling_comparing_2018,dorie_automated_2019}
% show benefits of outcome modeling to estimate average treatment effects.

%\idea{There is a variety of models for g-estimation}
% There is an explosion of outcome modeling or machine learning methods, either 
% developed for
% For instance, many deep-learning methods have been developed for medical
% image analysis \cite{shen2017deep,monshi2020deep}.
Outcome-modeling methods, even when specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand.
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models (as shown by \cite{dorie_automated_2019} and developed in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

Given complex,
potentially noisy, data, which model is to be most trusted to yield valid causal
estimates? As no single learner performs best on all data sets, there is a
pressing need for clear guidelines to select outcome models for causal
inference.

% TODO:move below
Standard practices to select models in predictive settings rely on
the error on the outcome
\cite{poldrack2020establishment,varoquaux2022evaluating}. However, as we
will see, these practices may not pick the best models
for causal inference, as they can be misled by inhomogeneities due to
treatment allocation.
%

@@

We study \textit{model selection procedures}
in practical settings: \textit{finite samples} settings and without
\textit{well-specification} assumption. Asymptotic causal-inference
theory calls for complex risks, but a practical question is
whether model-selection procedures, that rely on data split, can estimate
these risks reliably enough. Indeed, they
come with more quantities to estimate, which may
bring additional variance, leading to worse model selection.

We first illustrate the problem of causal model selection and briefly review
prior art. Then we anchors causal model selection in the \emph{potential
outcome} framework and details the causal risks and model-selection procedure.
We then rewrite the so-called $R\text{-risk}$, as a reweighted version of mean
squared difference between the true and estimated individualized treatment
effect. Finally, we conduct a thorough empirical study comparing the different
metrics on a wide variety of data-generative processes based on simulations and
real data.

% not the place for results
% Results outline how to best select outcome models for causal
% inference with an adapted
% cross-validation to estimate the so-called $R\text{-risk}$.
% This risk compensates for systematic
% differences between treated and non-treated individuals using
% two \emph{nuisance} models,
% themselves estimated from data and thus imperfect; yet these
% imperfections do not undermine the $R\text{-risk}$.
